---
title: "browsertime: live applink coldload, Chrome 74 vs GVE 68 vs Fenix 68 vs Fennec 64 on the Pixel 2 and Moto G5"
author: Nick Alexander
date: May 1, 2019
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    includes:
      in_header: "html/header.html"
---

<style>
select {
    width: 30%;
    right: 10px;
    height: 30px;
}
</style>

```{r setup, include=FALSE}
library(glue)
library(dplyr)
library(DT)
library(ggplot2)
library(stringr)
library(tidyr)
library(EnvStats) # For `geoMean`.

btr <- read.csv('data/applink.csv')
by_site <- btr %>%
  # filter(proxy == 'replay') %>%
  # select(-proxy) %>%
  select(-timestamp, -engine)
by_site$run = as.factor(by_site$run)
by_site$proxy = as.factor(by_site$proxy)
by_site$test_type = as.factor(by_site$test_type)

# We're not really interested in GVE: it measures Fenix-specific overhead, but GVE doesn't have TP enabled so it's not a direct comparison to Fenix.
# by_site <- by_site %>%
#  filter(vehicle != 'GVE 68')

# Ordering the vehicle factor in this way makes the dot-plot below generally "follow what we want": Fenix is fastest, Fennec is next, Chrome is slowest.
by_site$vehicle <- factor(by_site$vehicle, levels=c("Fenix 68",
                                                    "GVE 68",
                                                    "Fennec 64",
                                                    "Chrome 74"))

# Set up fixed color mapping for the different vehicles.  See https://stackoverflow.com/a/6920045.
library(RColorBrewer)
myColors <- brewer.pal(4, "Set1")[c(3, 2, 1, 4)] # Fenix=green, GVE=blue, Fennec=red, Chrome=purple.
names(myColors) <- levels(by_site$vehicle)
colScale <- scale_colour_manual(name = "vehicle", values = myColors)
fillScale <- scale_fill_manual(name = "vehicle", values = myColors)

shapeScale <- scale_shape_manual(name = "proxy", values=c(20, 18))

# Some sites have outliers that distort the scales.  Drop them for now.
bad_sites = c("bbc.") # co.uk")
for (bad_site in bad_sites) {
  by_site <- by_site %>%
    filter(!grepl(bad_site, site, fixed=TRUE));
}

runs <- data.frame(run = factor(c(1, 2, 3, 4)),
                   device = factor(c("Moto G5", "Moto G5", "Pixel 2", "Pixel 2")),
                   iterations = c(5, 5, 5, 5),
                   min_iterations = c(4, 4, 4, 4))

with_min_iterations <- by_site %>%
  group_by(run, device, site, vehicle) %>%
  summarise(num_iterations=n()) %>%
  ungroup() %>%
  group_by(run, device, site) %>%
  # This ensures that every "vehicle" has 0 if there were no runs recorded for that vehicle configuration.
  spread(vehicle, num_iterations, fill = 0) %>%
  gather("vehicle", num_iterations, levels(by_site$vehicle)) %>%
  ungroup() %>%
  group_by(run, device, site) %>%
  summarise(min_iterations=min(num_iterations)) %>%
  ungroup()

# Each (run, device, site) that has sufficient data to draw conclusions.
sufficient <- left_join(with_min_iterations, runs %>% select(-iterations), by = c("run", "device")) %>%
  filter(min_iterations.x >= min_iterations.y) %>%
  select(run, device, site)

insufficient <- left_join(with_min_iterations, runs %>% select(-iterations), by = c("run", "device")) %>%
  filter(min_iterations.x < min_iterations.y)

# The measurements corresponding to (run, device, site) tuples with sufficient data.
complete <- semi_join(by_site, sufficient, by = c("run", "device", "site"))

complete <- complete %>%
  group_by(device, run, site) %>%
  summarize(m=mean(pageLoadTime)) %>%
  ungroup() %>%
  arrange(m) %>%
  inner_join(complete, by=c('device', 'run', 'site'))

complete_sites <- complete %>% select(site) %>% distinct()
sites <- by_site %>% select(site) %>% distinct()
incomplete_sites <- setdiff(sites, complete_sites)

complete$site <- sub('https://', '', complete$site)
complete$site <- sub('http://', '', complete$site)
complete$site <- complete$site %>% str_trunc(30) %>% str_pad(30, 'right')
```

## tl;dr

This report summarizes data for the *applink* scenario: tapping a link in a third-party App (say, your mail client) and measuring the time that it takes Fenix to start, render the page, and fire the [`load` event](https://developer.mozilla.org/en-US/docs/Web/API/Window/load_event).  The times reported capture the time for the engine (GeckoView or Blink) to start and to complete pageload.

1. **A Quantum Flow-like startup performance effort to speed up
GeckoView startup is needed.** Chrome starts navigating in less than
0.5s, and so can we!

1. The particular site makes almost no difference to the engine
startup overhead , meaning that a few (or even one) representative
site(s) is sufficient to measure and track regressions in the applink
load time.  That will reduce the automation cost for regression
testing.

### Geometric means

The applink geometric means show that Fenix 68 takes roughly half as long as Fennec 64 to get to the `load` event, but takes more than one-and-a-half times as long as Chrome 74:

```{r, echo=FALSE}
geomeans <- complete %>%
  group_by(device, vehicle) %>%
  summarize(pageLoadTime=geoMean(pageLoadTime)) %>%
  ungroup()

xx <- geomeans %>% group_by(device) %>% spread(vehicle, pageLoadTime)

geomean_ratios <- xx %>% select(device)
geomean_ratios$f68_over_f64 <- xx$`Fenix 68` / xx$`Fennec 64`
geomean_ratios$f68_over_c74 <- xx$`Fenix 68` / xx$`Chrome 74`

datatable(arrange(geomean_ratios, device, f68_over_f64), options=list(dom='t', pageLength = nrow(geomean_ratios)),
          colnames = c('Fenix 68 over Fennec 64' = 'f68_over_f64',
                       'Fenix 68 over Chrome 74' = 'f68_over_c74'),
          fillContainer = FALSE,
          caption=('Applink pageload geomean ratios across sites (lower is better)' %>% glue())) %>%
  # formatRound(c("pageLoadTime"), 0)
  formatPercentage(c("Fenix 68 over Fennec 64")) %>%
  formatPercentage(c("Fenix 68 over Chrome 74"))
```

Here are the applink geomeans:

```{r, echo=FALSE}
datatable(arrange(geomeans, device, vehicle), options=list(dom='t', pageLength = nrow(geomeans)),
          colnames = c('pageLoadTime (ms)' = 'pageLoadTime'),
          fillContainer = FALSE,
          caption=('Applink pageload geomeans across sites (lower is better)' %>% glue())) %>%
  # formatRound(c("Fenix 68", "Chrome 74"), 2) %>%
  formatRound(c("pageLoadTime (ms)"), 0)
  # formatPercentage(c("pageLoadTime"))
```

## Methodology

### Test harness

The data were collected using an ad-hoc Python harness driving the
browsertime testing suite.  The harness is [still under
development](https://hg.mozilla.org/users/nalexander_mozilla.com/gecko/rev/face4e993f959724fe86b302987091666e3193cd)
but will eventually be published and development is tracked by [Bug
1545627](https://bugzilla.mozilla.org/show_bug.cgi?id=1545627).

browsertime drives the underlying vehicles using Web Driver
automation; for WebView this means `chromedriver` driving the engine
via the Chrome Debug Protocol and for GeckoView this means
`geckodriver` driving the engine over the Marionette protocol.

The version of browsertime used was lightly modified to support
Android-specific WebView engine configuration and to support the
GeckoView engine.  None of these modifications are believed to impact
engine performance.

The version of `geckodriver` was heavily modified to support the
GeckoView engine over the `adb` TCP/IP protocol.  These modifications
principally concern launching the target vehicle and connecting to the
underlying protocol handler; any impact on engine performance has to
do with servicing the underlying protocol and ambient engine
configuration (for example, custom profiles in GeckoView).

The version of `chromedriver` was manually patched to support
measuring applinks.  This required making two strings be empty in
order to change the `am start ...` invocation and in order to exploit
a further unsanitized string substitution.  A version of
`chromedriver` could have been compiled for this purpose, but the
Chromium build system is not easy to work with and this approach was
surprisingly simple.

### Vehicles tested

The data were collected from the following vehicle configurations:

| vehicle | engine | Tracking Protection |
| --- | --- | --- |
| Fenix 68 | GeckoView | enabled |
| geckoview_example 68 (GVE) | GeckoView | disabled |
| Fennec 64 | GeckoView | N/A |
| Chrome 74 | Blink | N/A |

### Single site test

For each site, the four vehicle configurations were tested using a
customized version of browsertime that starts the vehicle with an
Android Intent with `-a android.intent.action.VIEW -d $URL` and that
measures the process start time (in milliseconds after the epoch) and
the time that the engine reports starting the top-level navigation (in
milliseconds after the epoch).

The process start time in milliseconds after the epoch was determined
by inspecting `/proc/${PID}/stat` for the relevant process ID.  There
is some noise in this measurement (generally less than 10ms) so three
readings were averaged and rounded to the nearest millisecond.  The
calculations can be found (as browsertime modifications)
[here](https://github.com/ncalexan/browsertime/blob/4f8738e2225b3298556308c0d8c916d3b22e9741/lib/android/index.js#L100).

The engine's navigation start time origin is commensurable with the
process start time.

browsertime reports a wide range of timings, mostly from the
[Performance Navigation Timing
API](https://developer.mozilla.org/en-US/docs/Web/API/PerformanceNavigationTiming),
and the customized version of browsertime produces these as timings
relative to the process start time; see
[here](https://hg.mozilla.org/users/nalexander_mozilla.com/gecko/file/face4e993f959724fe86b302987091666e3193cd/testing/perftest/vendor/browsertime_applink.js#l16).

Schematically:

1. `am start ... -a android.intent.action.VIEW $URL`

1. `/proc/${PID}/stat` to get `processStartTime`

1. `browsertime` to get `window.performance.timeOrigin` and
`navigationStartTime`

1. Calculations to determine time to applink coldload:
`navigationStartTime` - `processStartTime` + `loadEventStart`

### Sites tested

The sites were taken from the [product mobile
corpus](https://docs.google.com/spreadsheets/d/1wqGfLaEKVDjfA-y4gZfcFRtjU3G0HzOxlcFTMlJGEJw/edit?ts=5bdb67e1#gid=596602279)
were tested.  The sites not tested were:

| site | reason |
| --- | --- |
| https://www.allrecipes.com/ | live load failures |
| https://www.allrecipes.com/recipe/16485/barbs-broccoli-cauliflower-salad/?internalSource=hub%20recipe&referringContentType=Search | live load failures |

Some sites witnessed transient network errors: in these cases the
number of recorded measurements is fewer than expected.  In any
individual run, no site was measured fewer than 4 times.

The entire corpus was tested end-to-end twice in succession on a
single Pixel 2 and a single Moto G5.

## Results

### Fenix 68 applink is significantly faster than Fennec 64

Fenix 68 applink is significantly faster than Fennec 64, but still lags Chrome.  Ignoring GVE, the following graph gives insight into how the vehicles compare on the sites in the test corpus for each run with live sites:

```{r, echo=FALSE, fig.width = 12, fig.height=12}
xx <- complete %>%
  filter(proxy == 'live') %>%
  filter(vehicle != 'GVE 68')
p <- ggplot(xx, aes(x=reorder(site, m), y=pageLoadTime)) + colScale
p <- p + geom_point(aes(color=vehicle), position=position_dodge(0.5))
p <- p + facet_wrap(vars(device), ncol=1, scales='fixed', labeller=labeller(.multi_line=FALSE)) + theme(axis.text.x=element_text(angle=-67.5, hjust=0, vjust=0)) + xlab('site')
p <- p + theme(legend.justification = c(0, 1), legend.position = c(0.04, 0.995))
p
```

### GeckoView takes longer to start than Chrome; all engines are consistent across sites

Within a particular vehicle on a particular device, the time from process start until the underlying engine reports that the applink top-level navigation has started is remarkably consistent across sites.  This time is essentially the "engine initialization time", and it's clear that Gecko is significantly slower than Chrome.

```{r, echo=FALSE, fig.width = 8, fig.height=8}

fun_mean <- function(x){
  return(data.frame(y=mean(x),label=round(mean(x,na.rm=T))))}

xx <- complete %>%
  filter(proxy == 'live') %>%
  group_by(run, device, vehicle)
p <- ggplot(xx, aes(x=vehicle, y=navigationStartTime)) + colScale
p <- p + geom_boxplot(aes(fill=vehicle, group=paste(as.integer(vehicle), run)), alpha=0.5) + fillScale
p <- p + geom_jitter(aes(color=vehicle, group=paste(as.integer(vehicle), run)), width=0.1, size=0.1)
p <- p + stat_summary(fun.data = fun_mean, geom="text", hjust=3.5, vjust=-3)
p <- p + facet_wrap(vars(device), ncol=1, scales='fixed', labeller=labeller(.multi_line=FALSE))
p <- p + theme(legend.justification = c(0, 1), legend.position = c(0.04, 0.995))
p
```

At this time, it is not clear why Fenix is *faster* than GVE to `navigationStartTime`.

### Fenix 68 vs Fennec 64: better than before

We see a nice win for Fenix 68 vs Fennec 64 across the board.

```{r, echo=FALSE, fig.width = 12, fig.height=12}
xx <- complete %>%
  filter(proxy == 'live') %>%
  filter(vehicle %in% c("Fenix 68", "Fennec 64"))

p <- ggplot(xx, aes(x=reorder(site, m), y=pageLoadTime)) + colScale
p <- p + geom_point(aes(colour=vehicle, group=paste(as.integer(vehicle), run)), position=position_dodge(0.5))
p <- p + facet_wrap(vars(device, proxy), ncol=1, scales='fixed', labeller=labeller(.multi_line=FALSE)) + theme(axis.text.x=element_text(angle=-67.5, hjust=0, vjust=0)) + xlab('site')
p <- p + theme(legend.justification = c(0, 1), legend.position = c(0.04, 0.995))
p
```

### Fenix 68 vs Chrome 74: losing today

Fenix 68 lags Chrome 74 across the board.  The startup overhead is so bad that the set of sites where tracking protection dramatically improves pageload time (as measured by firing the load event) is greatly reduced: it's only the worst offenders -- the buzzfeeds of the world -- where tracking protection can still shine.

```{r, echo=FALSE, fig.width = 12, fig.height=12}
xx <- complete %>%
  filter(proxy == 'live') %>%
  filter(vehicle %in% c("Fenix 68", "Chrome 74"))
p <- ggplot(xx, aes(x=reorder(site, m), y=pageLoadTime)) + colScale
p <- p + geom_point(aes(color=vehicle, group=paste(as.integer(vehicle), run)), position=position_dodge(0.5)) + shapeScale
p <- p + facet_wrap(vars(device, proxy), ncol=1, scales='fixed', labeller=labeller(.multi_line=FALSE)) + theme(axis.text.x=element_text(angle=-67.5, hjust=0, vjust=0)) + xlab('site')
p <- p + theme(legend.justification = c(0, 1), legend.position = c(0.04, 0.995))
p
```

### Fenix 68 vs GVE 68: the impact of tracking protection

Fenix 68 vs GVE 68 is a good proxy for the impact of tracking protection.

```{r, echo=FALSE, fig.width = 12, fig.height=12}
xx <- complete %>%
  filter(proxy == 'live') %>%
  filter(vehicle %in% c("Fenix 68", "GVE 68", "Chrome 74"))
p <- ggplot(xx, aes(x=reorder(site, m), y=pageLoadTime)) + colScale
p <- p + geom_point(aes(color=vehicle, group=paste(as.integer(vehicle), run)), position=position_dodge(0.5)) + shapeScale
p <- p + facet_wrap(vars(device, proxy), ncol=1, scales='fixed', labeller=labeller(.multi_line=FALSE)) + theme(axis.text.x=element_text(angle=-67.5, hjust=0, vjust=0)) + xlab('site')
p <- p + theme(legend.justification = c(0, 1), legend.position = c(0.04, 0.995))
p
```

## Inter-vehicle reliability

### Network weather and live site differences

No record and replay proxy was used to minimize the impact of network
weather.

Some of the sites serve dynamic content and/or advertisements.  This
means that between individual tests, the underlying network archives
may have changed significantly.

### Gecko profile conditioning

It is well known that the Gecko profile significantly impacts the
performance of the Gecko engine: preferences, certificate databases,
and the network cache itself can have major impacts on measurements.

To minimize volatility, for each GeckoView-based vehicle
configuration, a Gecko profile was conditioned as follows:

1. A profile template (with `cert9.db` and `key4.db` containing the
custom CA certificate used by proxies) was produced.

1. This profile template was copied to the target device, and the
vehicle was started from a cleared state with this profile.

1. The single page `https://example.com` was visited.

1. The vehicle was left idle for 2 minutes.

1. The vehicle was force-stopped and the conditioned profile
retrieved from the device.

This conditioned profile was then copied to the device at the
beginning of every test run: that is, every applink cold pageload
started with exactly the same Gecko profile.

The code that implements this profile conditioning can be found
[here](https://hg.mozilla.org/users/nalexander_mozilla.com/gecko/file/face4e993f959724fe86b302987091666e3193cd/testing/perftest/perftest/condition.py).

## Versions

### Device versions

The output of `adb shell getprop` for each device is available:

| device |
| --- |
| [Pixel 2](Pixel_2.getprop) |
| [Moto G5](Moto_G5.getprop) |

### Software versions

| package | version (* denotes modified) | link |
| ------- | --- | --- |
| browsertime | 4.7.0 (*) | https://github.com/ncalexan/browsertime/tarball/4f8738e2225b3298556308c0d8c916d3b22e9741 |
| chromedriver| 2.46 (*) | https://chromedriver.storage.googleapis.com/index.html?path=2.46/ |
| geckodriver | 0.40.0 (*) | https://hg.mozilla.org/users/nalexander_mozilla.com/gecko/rev/e7f1d26dec97a4a26c61b8e18dfe769d4c11e096 |
| Fenix | 1.0.1917 | https://queue.taskcluster.net/v1/task/TujyjjJqSoqeSVykLTjdmA/artifacts/public%2Ftarget.apk |
| geckoview_example | N/A | https://queue.taskcluster.net/v1/task/EbHvScwDQBWnoYHJRwZphQ/artifacts/public%2Fbuild%2Fgeckoview_example.apk |
| Fennec | 64.0.2 | http://archive.mozilla.org/pub/mobile/releases/64.0.2/android-api-16/multi/fennec-64.0.2.multi.android-arm.apk |
| Chrome | 74.0.3729.112 | N/A |

## Data

### Raw data

The data collected for the following tests is available:

| run | device | folder |
| --- | --- | --- |
| 1 | Moto G5 | https://drive.google.com/open?id=19tYjLirGCTWzyp75RxC3W68Ftf0rhdd0 |
| 2 | Moto G5 | https://drive.google.com/open?id=10mNf9yeqT_apryjjQg7bCEjdNPzcjV8s |
| 3 | Pixel 2 | https://drive.google.com/open?id=1EMYcH9eswZikgkhRpcm124__mWISkGZL |
| 4 | Pixel 2 | https://drive.google.com/open?id=1IZjT2stIJNyCIPVZhWKxXMXnYsehkIxw |

### Processed data

The data from the test runs above can be found in the following [CSV
file](data/applink.csv).  The columns of the CSV are as follows:

| column | description |
| --- | --- |
`device` | the target device, one of "Pixel 2", "Moto G5" |
| `run` | the test run number |
| `site` | the URL of the page being loaded |
| `engine` | the tested engine's User Agent string |
| `proxy` | "live" to signify the site was loaded from the live network |
| `timestamp` | the local timestamp when the pageload was initiated |
| `navigationStartTime` | the `window.performance.timeOrigin` reported by the engine under test, measured from the main process start time |
| `pageLoadTime` | the [`loadEventStart`](https://developer.mozilla.org/en-US/docs/Web/API/PerformanceNavigationTiming/loadEventStart) timestamp reported by the engine under test, measured from the main process start time |

### Fenix 68 vs Fennec 64 (in tabular form)

```{r, echo=FALSE}
f68_over_f64 <- complete %>%
  filter(vehicle %in% c("Fenix 68", "Fennec 64")) %>%
  group_by(run, device, site, vehicle) %>%
  summarise(pageLoadTime_mean=round(mean(pageLoadTime))) %>%
  ungroup() %>%
  group_by(run, device, site) %>%
  spread(vehicle, pageLoadTime_mean) %>%
  ungroup()

f68_over_f64$`ratio` = f68_over_f64$`Fenix 68` / f68_over_f64$`Fennec 64`

datatable(arrange(f68_over_f64, ratio), options=list(dom='t', pageLength = nrow(f68_over_f64)),
          fillContainer = FALSE,
          caption=('Fenix applink pageload over Fennec 64 applink pageload (lower is better)' %>% glue())) %>%
  # formatRound(c("Fenix 68", "Fennec 64"), 2) %>%
  formatPercentage(c("ratio"))
```

### Fenix 68 vs Chrome 74 (in tabular form)

```{r, echo=FALSE}
f68_over_c74 <- complete %>%
  filter(vehicle %in% c("Fenix 68", "Chrome 74")) %>%
  group_by(run, device, site, vehicle) %>%
  summarise(pageLoadTime_mean=round(mean(pageLoadTime))) %>%
  ungroup() %>%
  group_by(run, device, site) %>%
  spread(vehicle, pageLoadTime_mean) %>%
  ungroup()

f68_over_c74$`ratio` = f68_over_c74$`Fenix 68` / f68_over_c74$`Chrome 74`

datatable(arrange(f68_over_c74, ratio), options=list(dom='t', pageLength = nrow(f68_over_c74)),
          fillContainer = FALSE,
          caption=('Fenix applink pageload over Chrome applink pageload (lower is better)' %>% glue())) %>%
  # formatRound(c("Fenix 68", "Chrome 74"), 2) %>%
  formatPercentage(c("ratio"))
```
